### 2014-11-18 doc/thesis/system.tex
Adding text about the typing rules for the or idiom

The $\mathbf{or}$ idiom.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
\end{tabular}
\end{center}

Overloading on the input parameter.
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
& & \multicolumn{2}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; a \; {..} \; b$}\\
& & \multicolumn{2}{l}{$\mathbf{else}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; rep(a, b)_{e}$}\\
\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
\end{tabular}
\end{center}

@@

>>>>

Typed Lua includes four typing rules to handle the $\mathbf{or}$
logical operator and its common idioms:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-OR1}\\
\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
       \Nil \not\lesssim t \;\;\;
       \False \not\lesssim t}
      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-OR2}\\
\dfrac{\env_{1} \vdash e_{1}:\Nil, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3}}
      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-OR3}\\
\dfrac{\env_{1} \vdash e_{1}:\False, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3}}
      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:t, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-OR4}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
      {\env_{1} \vdash e_{1} \; \mathbf{or} \; e_{2}:filter(filter(t_{1}, \Nil), \False) \cup t_{2}, \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-OR1} is the rule that implements the short circuit.
We use the consistent-subtyping in this relation to guarantee that
the type system avaliates the second expression when the first one
has the dynamic type, as it can be hiding a false value.

The rules \textsc{T-OR2} and \textsc{T-OR3} guarantee that the final
result is the type of the second expression, because the first one
is certainly a false value.

The rule \textsc{T-OR4} is the most general rule, but it is also
the rule that handles the common $\mathbf{or}$ idioms.
It uses the predicate \emph{filter} to filter possible false values
that might be part of the type of the first expression.
We can use pattern matching to the define the recursive predicate
\emph{filter} as follows:
\begin{align*}
fu(t_{1} \cup t_{2}, t_{3}) & = fu(t_{1}, t_{3}) \cup fu(t_{2}, t_{3})\\
fu(t_{1} \cup t_{2}, t_{1}) & = fu(t_{2}, t_{1})\\
fu(t_{1} \cup t_{2}, t_{2}) & = fu(t_{1}, t_{2})\\
fu(t_{1}, t_{2}) & = t_{1}
\end{align*}

Using these typing rules our type system can type check the following
example:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
\end{tabular}
\end{center}

Without the \emph{filter} predicate,
the expression $x \; \mathbf{or} \; ``Hello"$ would have type
$\String \cup \Nil \cup ``Hello"$.
The \emph{filter} predicate removes the type $\Nil$ from the result,
leaving type $\String \cup ``Hello"$.
At the end of the evaluation, the expression has type $\String$
because unions are disjoint and $``Hello" \subtype \String$.

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
       closeall(\env_{1}[n \mapsto filter(\env_{1}(n), \String)) \vdash s_{2}:\env_{3}
      \end{array}}
      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}

\end{array}
\]

Using the rule \textsc{T-IF2}, Typed Lua can type check the following example:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
& & \multicolumn{2}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; a \; {..} \; b$}\\
& & \multicolumn{2}{l}{$\mathbf{else}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; rep(a, b)_{e}$}\\
\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
\end{tabular}
\end{center}

@@

Typed Lua also includes similar rules to handle the tags \texttt{nil},
\texttt{boolean}, and \texttt{number}.
